---
output:
  md_document:
    variant: markdown_github
---
[![Travis-CI Build Status](https://travis-ci.org/thierrygosselin/grur.svg?branch=master)](https://travis-ci.org/thierrygosselin/grur)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/thierrygosselin/grur?branch=master&svg=true)](https://ci.appveyor.com/project/thierrygosselin/grur)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/grur)](http://cran.r-project.org/package=grur)
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
[![DOI](https://zenodo.org/badge/14548/thierrygosselin/grur.svg)](https://zenodo.org/badge/latestdoi/14548/thierrygosselin/grur)

```{r, echo = FALSE}
description <- readLines("DESCRIPTION")
rvers <- stringr::str_match(grep("R \\(", description, value = TRUE), "[0-9]{1,4}\\.[0-9]{1,4}\\.[0-9]{1,4}")[1,1]
```
```{r, echo = FALSE}
description <- readLines("DESCRIPTION")
version <- gsub(" ", "", gsub("Version:", "", grep("Version:", description, value = TRUE)))
```
[![packageversion](https://img.shields.io/badge/Package%20version-`r version`-orange.svg)](commits/master)
[![Last-changedate](https://img.shields.io/badge/last%20change-`r gsub('-', '--', Sys.Date())`-brightgreen.svg)](/commits/master)

---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```


# grur: an R package tailored for RADseq data imputations

Numerous genomic analysis are vulnerable to missing values,
don't get trapped by missing genotypes in your RADseq dataset.
Use **grur** to **visualize patterns of missingness** and 
**perform map-independent imputations of missing genotypes** 
(see [features](https://github.com/thierrygosselin/grur#features) below).

## Installation
To try out the dev version of **grur**, copy/paste the code below:

```r
if (!require("devtools")) install.packages("devtools") # to install
devtools::install_github("thierrygosselin/grur")
library(grur) # to load
```

## Prerequisite - Suggestions - Troubleshooting
  * **Parallel computing**: Follow the steps in this [vignette](https://github.com/thierrygosselin/grur/blob/master/vignettes/vignette_imputations_parallel.Rmd) to install [randomForestSRC](http://www.ccs.miami.edu/~hishwaran/rfsrc.html) and [XGBoost](https://github.com/dmlc/xgboost) packages to really work in parallel.
  * **Installation problem:** see this
  [vignette](https://github.com/thierrygosselin/stackr/blob/master/vignettes/vignette_installation_problems.Rmd)
  * **Windows users**: Install [Rtools](https://cran.r-project.org/bin/windows/Rtools/).
  * For a better experience in **grur**, I recommend using [RStudio](https://www.rstudio.com/products/rstudio/download/). 
  The R GUI is unstable with functions using parallel ([more info](https://stat.ethz.ch/R-manual/R-devel/library/parallel/html/mclapply.html)). 

## Features
| Caracteristics | Description |
|:-------------------|:--------------------------------------------------------|
|**Patterns of missingness**|`missing_visualization`: visualize patterns of missing data associated with different variables of your study (lanes, chips, sequencers, populations, sample sites, reads/samples, homozygosity, etc). Similar to PLINK's identify-by-missingness analysis (IBM), **grur** is more powerful because it generates more analysis and automatically creates tables and figures.|
|**Imputations**|**Map-independent** imputations of missing genotypes with several algorithms:<br> * **Random Forests** (on-the-fly-imputations or using predictive modeling),<br>* **Extreme Gradient Tree Boosting**<br>* **Classic Strawman: ** the most frequently observed, non-missing, genotypes is used for imputation.<br><br>**Hierarchy: ** algorithm's model can account for *population* groupings or not (i.e. *overall samples*).<br><br>**Haplotype: ** automatically detect SNPs on the same LOCUS (read/haplotype) to impute the SNPs jointly, reducing imputation artifacts.|
| **Input/Output** | The imputations offered in **grur** are seamlesly integrated in **stackr** and **assigner**. Imputations are also integrated with usefull filters, blacklist and whitelist inside those 2 packages. Genetic formats recognized: [VCF, SNPs and haplotypes](https://samtools.github.io/hts-specs/), [PLINK tped/tfam](http://pngu.mgh.harvard.edu/~purcell/plink/data.shtml#tr), [genind](https://github.com/thibautjombart/adegenet), [genlight](https://github.com/thibautjombart/adegenet), [strataG gtypes](https://github.com/EricArcher/strataG), [Genepop](http://genepop.curtin.edu.au), [STACKS haplotype file](http://catchenlab.life.illinois.edu/stacks/), [hierfstat](https://github.com/jgx65/hierfstat), [COLONY](https://www.zsl.org/science/software/colony), [betadiv](http://adn.biol.umontreal.ca/~numericalecology/Rcode/), [δaδi](http://gutengroup.mcb.arizona.edu/software/), [structure](http://pritchardlab.stanford.edu/structure.html), [Arlequin](http://cmpg.unibe.ch/software/arlequin35/), [SNPRelate](https://github.com/zhengxwen/SNPRelate), dataframes of genotypes in wide or long/tidy format|
|**[ggplot2](http://ggplot2.org)-based plotting**|Visualization: publication-ready figures of important metrics and statistics.|
|**Parallel**|Codes designed and optimized for fast computations with progress bars. Works with all OS: Linux, Mac and yes PC!|

## Vignettes, R Notebooks and examples

Vignettes with real data for example in the form of R Notebooks take too much
space to be included in package, without CRAN complaining.
Consequently, vignettes are distributed separately, follow the links below.

* Missing data visualization and analysis [(html vignette)](https://www.dropbox.com/s/4zf032g6yjatj0a/vignette_missing_data_analysis.nb.html?dl=0) and [(Rmd)](https://www.dropbox.com/s/5fxw2h9w1l1j391/vignette_missing_data_analysis.Rmd?dl=0)


## Citation:
To get the citation, inside R:
```r
citation("grur")
```

## New features
Change log, version, new features and bug history lives in the [NEWS.md file](https://github.com/thierrygosselin/grur/blob/master/NEWS.md)

**v.0.0.1 2017-04-07**

* **grur** package launch!


## Roadmap of future developments:

* Integrate more imputation method.
* Workflow tutorial and vignettes to further explore some problems: *in progress*
* Use Shiny and ggvis (when subplots and/or facets becomes available for ggvis).
* Until publication **grur** will change rapidly, stay updated with releases and contribute with bug reports.
* Suggestions ?


## Contributions:

This package has been developed in the open, and it wouldn’t be nearly as good without your contributions. There are a number of ways you can help me make this package even better:

* If you don’t understand something, please let me know and raise an [issue](https://github.com/thierrygosselin/grur/issues)
* Your feedback on what is confusing or hard to understand is valuable. 
* If you spot a typo, feel free to edit the underlying page and send a pull request.

New to pull request on github ? The process is very easy:

* Click the edit this page on the sidebar.
* Make the changes using github’s in-page editor and save.
* Submit a pull request and include a brief description of your changes. 
* “Fixing typos” is perfectly adequate.
